---
title: "Interim Futility for Negative Binomial Endpoints"
author: "Sebastian Weber"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
  html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Interim Futility for Negative Binomial Endpoints}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, child="../settings-knitr.txt"}
```
```{r, child="../settings-sampling.txt"}
```
```{r init, include=FALSE}
library(dplyr)
library(MASS)

source(system.file("extra", "oc2S_interim.R", package = "RBesT",
                    mustWork = TRUE))
```

# Introduction

Clinical trials with recurrent event endpoints — such as asthma
exacerbations — typically require large sample sizes and long
follow-up. These features make early interim analyses (IAs) for
futility attractive: stopping a trial that is unlikely to succeed
saves resources and avoids exposing patients to ineffective
treatments.

However, a futility IA at a low information fraction (e.g.\ 30%) has
limited statistical power. The data collected so far carry substantial
uncertainty, making it difficult to distinguish a truly futile
scenario from random noise. This is where **historical borrowing**
via a Meta-Analytic-Predictive (MAP) prior can help: by incorporating
external evidence on the control rate and overdispersion, the MAP
prior effectively supplements the interim data, increasing the
information fraction available at the IA and enabling more precise
futility decisions.

This article develops the idea in four steps:

1. **Baseline:** Operating characteristics (OC) of a fixed-sample
   design without an IA.
2. **Futility IA without borrowing:** Introduce a futility stop at
   30% information fraction using non-informative priors. Quantify the
   power loss under the alternative.
3. **Futility IA with MAP prior:** Derive a MAP prior from four
   historical phase III trials. Show the effective sample size (ESS)
   and the information fraction gained.
4. **Interim decision-making:** Conditional power and predictive power
   (probability of success) at a hypothesized interim state.
5. **Design-level power:** Numerical integration over the full trial
   to quantify the expected power loss from the futility IA under the
   alternative.

Throughout, the **final confirmatory analysis is purely frequentist**
— the MAP prior is used only for interim projections.


# Negative binomial setup

We work on the **log mean rate** scale under the standard normal
approximation to the MLE (see [Appendix](#appendix) for the full
parametrization). The treatment effect is the log-rate ratio
$\delta = \log(\lambda_T / \lambda_C)$.

```{r nb-setup}
## Assumed true parameters for design planning
lambda_ctrl  <- 1.8          # placebo rate (events/patient-year)
kappa_true   <- 1.9          # overdispersion (phi)
followup     <- 1            # exposure per patient (years)
log_exposure <- log(followup)

## NB family object for OC calculations (fixed kappa)
nb_family <- negative.binomial(theta = 1 / kappa_true)

## Per-patient sampling SD on the log-rate scale
## sigma = sqrt((1 + kappa * lambda * t) / (lambda * t))
sigma_fn <- function(lambda, kappa, t) {
  sqrt((1 + kappa * lambda * t) / (lambda * t))
}
sigma_ctrl  <- round(sigma_fn(lambda_ctrl, kappa_true, followup), 2)
sigma_treat <- sigma_ctrl  # same rate assumed under H0

cat(sprintf("Per-patient sigma (log-rate scale): %.3f\n", sigma_ctrl))
```


# Fixed-sample design (no interim)

We first evaluate the operating characteristics of a conventional
2-arm trial without an interim analysis.

```{r fixed-design}
## Priors — non-informative for both arms (no borrowing)
uninf_ctrl  <- mixnorm(c(1, log(lambda_ctrl), 10), sigma = sigma_ctrl)
uninf_treat <- mixnorm(c(1, log(lambda_ctrl), 10), sigma = sigma_treat)

## Sample sizes
n_treat <- 150
n_ctrl  <- 150

## Decision: P(delta < 0 | data) > 0.975  (one-sided test)
success_crit <- decision2S(0.975, 0, lower.tail = TRUE)

## OC function
oc_fixed <- oc2S(uninf_treat, uninf_ctrl,
  n1 = n_treat, n2 = n_ctrl,
  decision = success_crit,
  family = nb_family,
  offset1 = log_exposure, offset2 = log_exposure
)
```

```{r fixed-power, fig.cap="Power curve — fixed-sample design (no IA)", echo = FALSE}
log_mu_ctrl <- log(lambda_ctrl)
log_rr <- seq(0, -0.8, by = -0.025)

power_fixed <- oc_fixed(log_mu_ctrl + log_rr,
                        rep(log_mu_ctrl, length(log_rr)))

power_df <- data.frame(log_rr = log_rr, power = power_fixed)

ggplot(power_df, aes(exp(log_rr), power)) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = 0.8, linetype = "dashed", colour = "grey50") +
  labs(x = "Rate ratio (treatment vs control)",
       y = "Power",
       title = "Power curve — no interim analysis") +
  scale_x_log10()
```


# Futility IA at 30% information fraction (no borrowing)

We now add a futility stop at 30% of the planned information. With no
historical borrowing, the IA relies entirely on the accrued trial
data. This would be a very early stop with rather imprecise estimates
compared to a more common IA look at 50%.

The information fraction is defined as the ratio of the Fisher
information for the treatment effect (log rate ratio) at the interim
relative to the final analysis. Under a negative binomial model with
log link, the per-patient Fisher information for $\log\mu$ in arm $j$
is $I_j = \lambda_j t / (1 + \kappa\lambda_j t)$, and the information
for the treatment effect $\delta = \log\mu_T - \log\mu_C$ from $n_T$
and $n_C$ patients is
$$
\mathcal{I}(\delta) = \frac{1}{(n_T I_T)^{-1} + (n_C I_C)^{-1}}.
$$
We compute the interim sample sizes required for 30\% of
$\mathcal{I}(\delta)$ at the planned final sample size, evaluated
under the design alternative $\log(\text{RR}) = -0.5$:

```{r ia-setup, echo=FALSE}
## Design alternative: log rate ratio at which trial is powered (~80%)
log_rr_design        <- -0.5
lambda_treat_design  <- lambda_ctrl * exp(log_rr_design)

## Per-patient Fisher information for log(mu) under each arm
## In a NB GLM with log link: I_i = lambda*t / (1 + kappa*lambda*t)
fisher_per_patient <- function(lambda, kappa, t) {
  lambda * t / (1 + kappa * lambda * t)
}

Ip_ctrl  <- fisher_per_patient(lambda_ctrl, kappa_true, followup)
Ip_treat <- fisher_per_patient(lambda_treat_design, kappa_true, followup)

## Fisher information for delta at planned sample size
info_final  <- 1 / (1 / (n_treat * Ip_treat) + 1 / (n_ctrl * Ip_ctrl))

## Target: 30% of final information at the IA
info_frac_ia <- 0.30
info_target  <- info_frac_ia * info_final

## Solve for interim n assuming 1:1 allocation preserved
n_ia_exact  <- info_target * (1 / Ip_treat + 1 / Ip_ctrl)
n_ctrl_ia   <- round(n_ia_exact)
n_treat_ia  <- round(n_ia_exact)

knitr::kable(data.frame(
  Quantity = c("Design alternative log(RR)",
               "Per-patient Fisher info (treatment)",
               "Per-patient Fisher info (control)",
               "Information at final (n/arm)",
               sprintf("Information target at IA (%.0f%%)", 100 * info_frac_ia),
               "Patients at IA (treatment / control)",
               "Patients remaining (treatment / control)"),
  Value = c(sprintf("%.2f", log_rr_design),
            sprintf("%.4f", Ip_treat),
            sprintf("%.4f", Ip_ctrl),
            sprintf("%.2f (n = %d)", info_final, n_ctrl),
            sprintf("%.2f", info_target),
            sprintf("%d / %d", n_treat_ia, n_ctrl_ia),
            sprintf("%d / %d", n_treat - n_treat_ia, n_ctrl - n_ctrl_ia))
))
```

The futility rule is based on **predictive power (probability of
success)**: given the interim posterior, what is the probability that
the final analysis — using all planned patients — will declare
success?

If the PoS falls below a futility threshold $\gamma_F$ (e.g.\ 10%),
we stop:

```{r futility-rule}
## Futility threshold for PoS
post_futility_thresh <- 0.10
```

## Operating characteristics of the IA design

To evaluate the operating characteristics of the design *with* the
futility IA, we use numerical integration via `oc2S_interim()`.  This
function avoids noisy Monte Carlo estimation via Gauss-Hermite
quadrature over the distribution of possible interim outcomes (see
[Appendix: Integration approach](#appendix-integration) for details).

The interim decision rule continues the trial whenever the
probability of success exceeds the futility threshold:

```{r oc-ia-noprior, fig.cap="Power with and without futility IA (no borrowing)"}
## IA continuation rule: continue if PoS > threshold
## Note: we use the fixed-sigma path of pos2S (no family argument)
## for performance. This is valid because sigma is approximately
## constant over the posterior range at the interim.
ia_rule_uninf <- function(post1_ia, post2_ia) {
  pos2S(post1_ia, post2_ia,
        n1 = n_treat - n_treat_ia, n2 = n_ctrl - n_ctrl_ia,
        decision = success_crit,
        sigma1 = sigma_ctrl, sigma2 = sigma_ctrl)(post1_ia, post2_ia) > post_futility_thresh
}

## OC function with futility IA (no borrowing)
oc_ia_uninf <- oc2S_interim(
  prior1 = uninf_treat, prior2 = uninf_ctrl,
  n1 = n_treat, n2 = n_ctrl,
  n1_ia = n_treat_ia, n2_ia = n_ctrl_ia,
  decision = success_crit,
  ia_rule = ia_rule_uninf,
  family = nb_family,
  offset1 = log_exposure, offset2 = log_exposure,
  Ngrid_ia = 11L
)

## Evaluate at null and design alternative (pos2S rule is expensive)
log_rr_grid <- c(`No benefit` = 0, `Design alt` = -0.6)

power_ia_uninf_df <- oc_ia_uninf(
  theta1 = log_mu_ctrl + log_rr_grid,
  theta2 = rep(log_mu_ctrl, length(log_rr_grid))
)
power_ia_uninf_df$log_rr <- log_rr_grid 
```

```{r power-comparison-noprior}
## Compare with fixed-design power at same grid points
power_fixed_grid <- oc_fixed(log_mu_ctrl + log_rr_grid,
                             rep(log_mu_ctrl, length(log_rr_grid)))
```

```{r power-comparison-noprior-tab, echo=FALSE}
knitr::kable(data.frame(
  Scenario = names(log_rr_grid),
  `Rate ratio` = round(exp(log_rr_grid), 2),
  `Fixed design` = power_fixed_grid,
  `IA (no borrow)` = power_ia_uninf_df$power,
  `Power loss` = power_fixed_grid - power_ia_uninf_df$power,
  `Futility stop` = power_ia_uninf_df$stop_prob,
  check.names = FALSE
), digits = 3, row.names = FALSE,
caption = "Power and futility-stop probability by scenario (no borrowing, PoS rule)")
```

At 30% information fraction with non-informative priors, the futility
IA causes a noticeable power loss under the alternative while the
futility stopping probability under the null is moderate. The interim
data simply do not carry enough information for reliable futility
decisions.


# MAP prior from historical phase III trials

## Historical data

We use placebo-arm data from the phase III asthma trials in the
`asthma` data set (Table 1 of Holzhauer, Wang & Schmidli, 2018):

```{r hist-data}
asthma_ph3 <- subset(asthma, phase == "phase III")
```

```{r hist-data-tab, echo=FALSE}
knitr::kable(asthma_ph3[, c("study", "d", "n", "log_mu_hat", "se_log_mu_hat", "kappa_hat")],
             digits = 2,
             caption = "Historical phase III placebo arms")
```

## MAP prior for the control log-rate

The historical trials have different exposure (follow-up) durations
$d$. Since the model is on the log-rate scale, we include
$\log(d)$ as an **offset** in `gMAP` so that the estimated parameter
is the log-rate $\log\lambda_C$ rather than the log-mean count
$\log(\mu) = \log(\lambda \cdot d)$:

```{r map-rate}
map_mcmc <- gMAP(
  cbind(log_mu_hat, se_log_mu_hat) ~ 1 | study,
  offset = log(d),
  data   = asthma_ph3,
  family = gaussian,
  tau.dist  = "HalfNormal",
  tau.prior = sigma_ctrl / 4,
  beta.prior = 2
)

print(map_mcmc)
map_rate <- automixfit(map_mcmc)
sigma(map_rate) <- sigma_ctrl
```

```{r map-forest, fig.cap="Forest plot of MAP model for control log-rate"}
pl <- plot(map_mcmc)
print(pl$forest_model)
```

## Effective sample size and information fraction gain

The ESS tells us how many patients the MAP prior is worth. It adds
precision to the **control arm only**, so we fold it into the
treatment-effect information $\mathcal{I}(\delta)$ defined above as
$\text{ESS}$ extra control patients. The information fraction at the IA
then uses the same $\mathcal{I}(\delta)/\mathcal{I}(\delta)_{\text{final}}$
ratio, now with $n_C \to n_{C,\text{IA}} + \text{ESS}$:

```{r ess-info}
ess_map <- round(ess(map_rate, sigma = sigma_ctrl))

## Effect information I(delta) at IA without borrowing
info_ia_no_borrow <- 1 / (1 / (n_treat_ia * Ip_treat) +
                          1 / (n_ctrl_ia * Ip_ctrl))
info_frac_no_borrow <- info_ia_no_borrow / info_final

## Effect information I(delta) at IA with MAP prior:
## ESS adds patient-equivalents to the CONTROL arm only.
info_ia_with_map <- 1 / (1 / (n_treat_ia * Ip_treat) +
                         1 / ((n_ctrl_ia + ess_map) * Ip_ctrl))
info_frac_with_map <- info_ia_with_map / info_final
```

```{r ess-info-tab, echo=FALSE}
knitr::kable(data.frame(
  Quantity = c("MAP prior ESS",
               "Info fraction at IA (no borrowing)",
               "Info fraction at IA (with MAP prior)",
               "Info fraction gained"),
  Value = c(sprintf("%d patients", ess_map),
            sprintf("%.1f%%", 100 * info_frac_no_borrow),
            sprintf("%.1f%%", 100 * info_frac_with_map),
            sprintf("+%.1f pp", 100 * (info_frac_with_map - info_frac_no_borrow)))
))
```

The MAP prior substantially increases the effective information
fraction at the IA, making futility decisions more reliable.

## Power with futility IA and MAP prior

```{r oc-ia-map}
## IA continuation rule with MAP prior on control for prediction.
## pos2S boundary uses the uninformative posteriors (final analysis
## is uninformative), but we integrate out future uncertainty using
## the MAP-informed posterior for control (post2_ia_info).
ia_rule_map <- function(post1_ia, post2_ia, post1_ia_info, post2_ia_info) {
  pos2S(post1_ia, post2_ia,
        n1 = n_treat - n_treat_ia, n2 = n_ctrl - n_ctrl_ia,
        decision = success_crit,
        sigma1 = sigma_ctrl, sigma2 = sigma_ctrl)(post1_ia, post2_ia_info) > post_futility_thresh
}

## OC function with futility IA and MAP prior
## prior1/prior2 are uninformative (used for the final analysis boundary).
## prior2_ia is the MAP prior (used only for the IA decision).
oc_ia_map <- oc2S_interim(
  prior1 = uninf_treat, prior2 = uninf_ctrl,
  n1 = n_treat, n2 = n_ctrl,
  n1_ia = n_treat_ia, n2_ia = n_ctrl_ia,
  decision = success_crit,
  ia_rule = ia_rule_map,
  prior2_ia = map_rate,
  family = nb_family,
  offset1 = log_exposure, offset2 = log_exposure,
  Ngrid_ia = 11L
)

## Evaluate at two scenarios: no benefit (rate ratio = 1) and the
## design alternative. The pos2S-based rule is expensive, so we keep
## the grid small.
scenarios   <- c(`No benefit` = 0, `Design alt` = log_rr_design_alt <- -0.6)
theta1_eval <- log_mu_ctrl + scenarios
theta2_eval <- rep(log_mu_ctrl, length(scenarios))

map_grid    <- oc_ia_map(theta1 = theta1_eval, theta2 = theta2_eval)
uninf_grid  <- oc_ia_uninf(theta1 = theta1_eval, theta2 = theta2_eval)
fixed_power <- oc_fixed(theta1_eval, theta2_eval)
```

```{r oc-ia-map-tab, echo=FALSE}
knitr::kable(data.frame(
  Scenario = names(scenarios),
  `Rate ratio` = round(exp(scenarios), 2),
  `Power (no IA)` = fixed_power,
  `Power (no borrow)` = uninf_grid$power,
  `Power (MAP)` = map_grid$power,
  `Stop (no borrow)` = uninf_grid$stop_prob,
  `Stop (MAP)` = map_grid$stop_prob,
  check.names = FALSE
), digits = 3, row.names = FALSE,
caption = "Power and futility-stop probability by scenario")
```

The table covers two complementary scenarios. A noticable result
appears under the **design alternative**: the MAP prior almost
eliminates the power loss caused by the futility IA. Without borrowing,
inserting the IA costs about 2.4 percentage points of power
(88.9% → 86.5%); with the MAP prior the loss shrinks to roughly 0.6
percentage points (88.9% → 88.3%), recovering about three quarters of
the lost power and landing essentially back at the fixed-design value.
This is remarkable: borrowing historical control information at the
interim makes the futility rule almost "free" under the alternative,
because the better-informed interim posterior leads to far fewer
erroneous futility stops (stop probability 4.6% → 1.3%). Under **no
treatment benefit** (rate ratio = 1), the relevant quantity is instead
the futility-stop probability: a useful IA should stop early with high
probability when there is nothing to gain. Both designs stop with high
probability here, so the IA is doing its job, while power under the
design alternative is preserved.

It is worth noting the **trade-off** behind these numbers. Borrowing
historical control information raises the interim information fraction
(30% → ~44%), which makes the control estimate more precise and the
futility decision less reactive to noise. Under the null the MAP stop
probability is actually *lower* (57% vs 63%): without borrowing, the
noisy 30%-information control estimate pushes more interim outcomes
across the futility threshold — some for the wrong reason. With the
MAP prior the predictive probability of success clusters nearer its
true value, so fewer outcomes are stopped by chance. The same noise
reduction is what protects power under the alternative. Borrowing
therefore buys sharper, less reactive decisions — not simply more
stopping.

The intro section deliberately powers the IA-rule grid at just these
two anchor scenarios for speed; evaluating `oc_ia_map` / `oc_ia_uninf`
over a finer `log_rr` grid traces the full power and stop-probability
curves at higher computational cost.


# Conditional and predictive power at interim

At an actual IA, two complementary quantities inform the go/no-go
decision:

- **Conditional power (CP):** The probability of final success
  *given a specific assumed true treatment effect*. This is a
  frequentist concept — the unknown parameters are fixed.
- **Predictive power / Probability of Success (PoS):** The
  probability of final success *integrating over parameter
  uncertainty*. This is a Bayesian concept — the unknown parameters
  are drawn from the interim posterior.

## Hypothesized interim state

We consider a scenario where at the IA (30% information fraction),
the observed data suggest a moderate treatment effect:

```{r interim-state}
## Observed interim summary statistics
ia_log_mu_ctrl  <- 0.70     # placebo MLE on log-rate scale
ia_se_mu_ctrl   <- sigma_fn(exp(0.70), kappa_true, followup) / sqrt(n_ctrl_ia)
ia_log_mu_treat <- 0.35     # treatment MLE (log RR ≈ -0.35)
ia_se_mu_treat  <- sigma_fn(exp(0.35), kappa_true, followup) / sqrt(n_treat_ia)
```

```{r interim-state-tab, echo=FALSE}
knitr::kable(data.frame(
  Quantity = c("Interim log-rate control",
               "Interim log-rate treatment",
               "Observed log RR"),
  Value = c(sprintf("%.2f", ia_log_mu_ctrl),
            sprintf("%.2f", ia_log_mu_treat),
            sprintf("%.2f", ia_log_mu_treat - ia_log_mu_ctrl)),
  SE = c(sprintf("%.3f", ia_se_mu_ctrl),
         sprintf("%.3f", ia_se_mu_treat),
         "")
))
```

## Interim posteriors

```{r interim-posteriors}
## Control posterior with MAP prior
post_ctrl_ia <- postmix(map_rate, m = ia_log_mu_ctrl, se = ia_se_mu_ctrl)

## Treatment posterior (non-informative prior)
post_treat_ia <- postmix(uninf_treat, m = ia_log_mu_treat, se = ia_se_mu_treat)
```

## Conditional power

Conditional power evaluates the probability of final success
*assuming the true treatment effect equals a specific value*.
We construct this using `oc2S` with the interim posteriors as
"priors" for the remaining data:

```{r conditional-power, fig.cap="Conditional power at interim"}
## OC function conditioned on interim posteriors
cp_fn <- oc2S(post_treat_ia, post_ctrl_ia,
  n1 = n_treat - n_treat_ia,
  n2 = n_ctrl  - n_ctrl_ia,
  decision = success_crit,
  family = nb_family,
  offset1 = log_exposure, offset2 = log_exposure
)

## Evaluate across true log-rate ratios
log_rr_cp <- seq(0, -0.8, by = -0.025)
cp_values <- cp_fn(log_mu_ctrl + log_rr_cp,
                   rep(log_mu_ctrl, length(log_rr_cp)))

ggplot(data.frame(log_rr = log_rr_cp, cp = cp_values),
       aes(exp(log_rr), cp)) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = c(0.5, 0.8),
             linetype = "dashed", colour = "grey50") +
  geom_vline(xintercept = exp(ia_log_mu_treat - ia_log_mu_ctrl),
             linetype = "dotted", colour = "steelblue") +
  annotate("text",
           x = exp(ia_log_mu_treat - ia_log_mu_ctrl) * 1.05,
           y = 0.95, label = "observed effect",
           hjust = 0, size = 3, colour = "steelblue") +
  labs(x = "Assumed true rate ratio",
       y = "Conditional power",
       title = "Conditional power at 30% IA") +
  scale_x_log10()
```

## Predictive power (Probability of Success)

Predictive power integrates over the interim posterior distribution of
the treatment effect, providing a single summary that accounts for
parameter uncertainty:

```{r predictive-power}
## PoS for remaining patients given interim posteriors
pos_fn <- pos2S(post_treat_ia, post_ctrl_ia,
  n1 = n_treat - n_treat_ia,
  n2 = n_ctrl  - n_ctrl_ia,
  decision = success_crit,
  family = nb_family,
  offset1 = log_exposure, offset2 = log_exposure
)

## Compute predictive power and conditional power at observed effect in one pass
pos_value <- pos_fn(post_treat_ia, post_ctrl_ia)
cp_at_obs <- cp_fn(ia_log_mu_treat, ia_log_mu_ctrl)
```

```{r predictive-power-tab, echo=FALSE}
knitr::kable(data.frame(
  Metric = c("Conditional power at observed effect",
             "Predictive power (PoS)"),
  Value = c(sprintf("%.3f", cp_at_obs),
            sprintf("%.3f", pos_value))
))
```

The predictive power is typically lower than the conditional power
evaluated at the observed effect, because it accounts for the
possibility that the true effect is smaller (or absent).


# Design-level power with futility IA

The individual-scenario analysis above is useful at the IA itself.
At the **design stage**, we want to know the overall impact of the
futility IA on power — averaging over all possible interim outcomes
under an assumed truth.

For Normal-endpoint group-sequential Bayesian designs, the
[gsbDesign](https://cran.r-project.org/package=gsbDesign) package
(Gerber & Gsponer, 2016) provides a full analytic treatment of such
operating characteristics via conjugate posterior updates. The
`oc2S_interim()` helper used here follows the same numerical
integration idea but works with the mixture-prior machinery in RBesT
(see [Appendix: Integration approach](#appendix-integration)).
We already computed these above. The table at the design alternative
(rate ratio ≈ 0.55) confirms that the MAP prior reduces the power
loss from the futility IA while increasing the futility stopping
probability under the null — exactly the desired behavior.


# Discussion

This article demonstrated a simple progression for using historical
borrowing in the design of a futility interim analysis:

1. **Establish the baseline** without an IA.
2. **Quantify the cost** of adding a futility IA with limited data.
3. **Show the benefit** of a MAP prior, both in ESS/information
   fraction terms and in operating characteristics.
4. **Provide interim decision tools** via conditional and predictive
   power.
5. **Confirm at the design level** that the power loss is acceptable
   via numerical integration.

Key insights:

- A futility IA at low information fractions (e.g.\ 30%) is unreliable
  without supplementary information. The uncertainty in the interim
  estimate is too large.
- The MAP prior adds effective sample size to the control arm, boosting
  the information fraction at the IA and enabling sharper futility
  decisions.
- The final confirmatory analysis remains frequentist — no prior is
  imported. The MAP prior serves only as a **calibration tool** for
  interim projections.
- Predictive power (PoS) is the more conservative and realistic
  interim metric compared to conditional power, because it integrates
  over parameter uncertainty.

It is left as an outlook to make use of the historical information of
the overdisperion parameter, which can be synthesized in a
normal-normal hierarchical model and also used as a MAP prior at the
IA. However, a random overdispersion parameter is not supported with
`RBesT` directly (assuming a known overdispersion is a common
assumption at the design stage).

## Limitations

- Overdispersion $\kappa$ is treated as fixed. A more thorough
  analysis would derive a MAP prior for $\kappa$ and integrate over
  its posterior uncertainty (cf.\ Holzhauer et al., 2018; see the
  [companion article on SSR](introduction_negbin_ia_ssr.html)).
- The normal approximation on the log-rate scale assumes moderate to
  large event counts. For rare-event settings, exact likelihoods may
  be needed.

### References

[1] Neuenschwander B. et al., *Clin Trials*. 2010; 7(1):5-18  
[2] Holzhauer B., Wang C., Schmidli H. Evidence synthesis from
    aggregate recurrent event data for clinical trial design and
    analysis. *Statistics in Medicine*. 2018;37:867-882.  
[3] Schmidli H. et al., *Biometrics* 2014;70(4):1023-1032  
[4] Gerber F, Gsponer T (2016). "gsbDesign: An R Package for Evaluating
    the Operating Characteristics of a Group Sequential Bayesian Design."
    *Journal of Statistical Software*, **69**(11), 1–23.
    doi: [10.18637/jss.v069.i11](https://doi.org/10.18637/jss.v069.i11).  
[5] Gsponer T, Gerber F, Bornkamp B, Ohlssen D, Vandemeulebroecke M,
    Schmidli H (2014). "A Practical Guide to Bayesian Group Sequential
    Designs." *Pharmaceutical Statistics*, **13**(1), 71–80.
    doi: [10.1002/pst.1593](https://doi.org/10.1002/pst.1593).  

### R Session Info

```{r session, echo=FALSE}
sessionInfo()
```

```{r, include=FALSE}
options(.user_mc_options)
```

# Appendix: Integration approach for interim OC {#appendix-integration}

Computing the operating characteristics of a design with an interim
analysis requires averaging over all possible interim outcomes.
The `oc2S_interim()` function (available via
`system.file("extra", "oc2S_interim.R", package = "RBesT")`) does
this via **deterministic numerical integration** rather than Monte
Carlo simulation, yielding smooth, noise-free OC curves.

## Decomposition

The unconditional power is

$$
P(\text{success}) = E_{\bar{y}_{IA}}
\Big[\mathbf{1}\{\text{continue}(\bar{y}_{IA})\}
\cdot P(\text{success} \mid \bar{y}_{IA}, \theta)\Big]
$$

where $\bar{y}_{IA} = (\bar{y}_{1,IA}, \bar{y}_{2,IA})$ are the
interim summary statistics (MLEs on the log-rate scale) and $\theta =
(\theta_1, \theta_2)$ are the true parameters.

The outer expectation is a 2D integral over the joint distribution of
the interim summaries $\bar{y}_{IA} \sim N(\theta, \Sigma_{IA})$.
Since the two arms are independent, this factors into a product of
two 1D Normal integrals — handled by **Gauss-Hermite quadrature**.

## Full-trial boundary reuse

The key performance insight is that the **full-trial decision
boundary** $B(\bar{y}_2)$ can be computed once and reused at every
quadrature point.

The final analysis decides based on the full-sample means
$\bar{y}_{\text{full}}$, which are a deterministic function of the
interim and remaining data:

$$
\bar{y}_{j,\text{full}} =
\frac{n_{j,IA}\,\bar{y}_{j,IA} + n_{j,rem}\,\bar{y}_{j,rem}}{n_j}
$$

Conditional on the interim data $\bar{y}_{j,IA}$ and the true
parameter $\theta_j$, the full-sample mean is Normal:

$$
\bar{y}_{j,\text{full}} \mid \bar{y}_{j,IA},\theta_j \sim
N\!\left(
  \frac{n_{j,IA}\,\bar{y}_{j,IA} + n_{j,rem}\,\theta_j}{n_j},\;
  \frac{\sigma_j \sqrt{n_{j,rem}}}{n_j}
\right)
$$

The decision boundary $B(\bar{y}_{2,\text{full}})$ — computed by
`decision2S_boundary()` using the original priors and full sample
sizes — applies unchanged, because the sequential Bayesian update
property guarantees

$$
p(\theta \mid \bar{y}_{\text{full}})
= p(\theta \mid \bar{y}_{IA}, \bar{y}_{rem}).
$$

At each quadrature point, the conditional OC reduces to

$$
P(\text{success} \mid \bar{y}_{IA}, \theta) =
\int \Phi\!\Big(B(\bar{y}_{2,\text{full}}),\;
\mu_{1,\text{cond}},\; \text{se}_{1,\text{cond}}\Big)\;
\varphi(\bar{y}_{2,\text{full}};\,
\mu_{2,\text{cond}},\,\text{se}_{2,\text{cond}})\;
d\bar{y}_{2,\text{full}}
$$

which is again evaluated by Gauss-Hermite quadrature (an inner 1D integral).


# Appendix: Negative binomial parametrization {#appendix}

## Distribution

For patient $k$ in arm $j$ with follow-up (exposure) time $t_{jk}$
and mean event rate $\lambda_j$ (events per unit time), the count
$Y_{jk}$ follows

$$Y_{jk} \sim \text{NB}(\mu_{jk} = t_{jk}\,\lambda_j,\; \kappa)$$

with probability mass function

$$P(Y_{jk} = y) =
  \binom{y + \kappa^{-1} - 1}{y}\,
  \left(\frac{\kappa\,\mu_{jk}}{1 + \kappa\,\mu_{jk}}\right)^y\,
  \left(\frac{1}{1 + \kappa\,\mu_{jk}}\right)^{\kappa^{-1}}$$

and moments

$$E(Y_{jk}) = \mu_{jk} = t_{jk}\,\lambda_j, \qquad
  \text{Var}(Y_{jk}) = \mu_{jk} + \kappa\,\mu_{jk}^2
  = \mu_{jk}(1 + \kappa\,\mu_{jk}).$$

Here $\kappa \ge 0$ is the **overdispersion** parameter. For
$\kappa = 0$ the distribution reduces to Poisson. The treatment
effect is parametrized as the **rate ratio**
$\theta = \lambda_T/\lambda_C$, or equivalently the **log-rate ratio**
$\delta = \log\theta = \log\lambda_T - \log\lambda_C$.

## Parametrization conventions

Several conventions coexist in the literature:

| Convention | Overdispersion | Relation |
|------------|---------------|----------|
| $\kappa$ (this article, Holzhauer et al.) | $\kappa \ge 0$ | — |
| $\phi$ (Mütze et al., rpact) | $\phi \ge 0$ | $\phi = \kappa$ |
| `theta` in `MASS::negative.binomial()` | `theta` $> 0$ | `theta` $= 1/\kappa$ |
| `size` in `dnbinom()` | `size` $> 0$ | `size` $= 1/\kappa$ |

**Caution:** R's `MASS::negative.binomial(theta)` uses `theta` as
the *reciprocal* of the overdispersion ($1/\kappa$), so larger values
of `theta` mean *less* overdispersion.

## Fisher information on the log-rate scale

The analysis is conducted on the **log mean rate** scale,
$\nu_j = \log\lambda_j$. The MLE $\hat\nu_j$ is approximately
normally distributed with variance (Fisher information inverse)

$$\text{Var}(\hat\nu_j) \approx
  \frac{1 + \kappa\,\lambda_j\,t}{n_j\,\lambda_j\,t}$$

for $n_j$ patients each with exposure $t$. The approximate Fisher
information for arm $j$ is therefore

$$\mathcal{I}_j = \frac{n_j\,\lambda_j\,t}{1 + \kappa\,\lambda_j\,t}.$$

Key observations:

- **Precision is event-driven:** The numerator $n_j\lambda_j t$
  represents the expected total number of events. More events yield
  more information.
- **Overdispersion is a tax:** The denominator
  $1 + \kappa\lambda_j t$ inflates the variance. Higher $\kappa$
  reduces the information per patient.
- **Exposure matters:** Longer follow-up $t$ increases the expected
  events and thus the information, but with diminishing returns when
  $\kappa\lambda_j t \gg 1$.

The variance of the log-rate ratio estimator is

$$\text{Var}(\hat\delta) \approx
  \frac{1}{\mathcal{I}_T} + \frac{1}{\mathcal{I}_C}.$$

## Connecting to `MASS::negative.binomial()`

In RBesT's design functions (`oc1S`, `oc2S`, `pos1S`, `pos2S`), the
`family` argument accepts an object created by
`MASS::negative.binomial()`. This object encodes the variance–mean
relationship $V(\mu) = \mu + \mu^2/\texttt{theta}$, which on the
log-link scale translates exactly to the Fisher information formula
above.

```{r family-example}
## Create NB family with kappa = 1.9 (theta = 1/1.9)
kappa <- 1.9
nb <- MASS::negative.binomial(theta = 1 / kappa)

## The variance function: V(mu) = mu + mu^2 / theta = mu * (1 + kappa * mu)
mu_grid <- seq(0.5, 3, by = 0.5)
data.frame(
  mu     = mu_grid,
  V_mu   = nb$variance(mu_grid),
  check  = mu_grid * (1 + kappa * mu_grid)
)
```

```{r,include=FALSE}
options(.user_mc_options)
```
