---
title: "Interim Futility for Negative Binomial Endpoints"
author: "Sebastian Weber"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
  html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Interim Futility for Negative Binomial Endpoints}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, child="../settings-knitr.txt"}
```
```{r, child="../settings-sampling.txt"}
```
```{r init, include=FALSE}
library(dplyr)
library(MASS)

source(system.file("extra", "oc2S_interim.R", package = "RBesT",
                    mustWork = TRUE))
```

# Introduction

Clinical trials with recurrent event endpoints — such as asthma
exacerbations — typically require large sample sizes and long
follow-up. These features make early interim analyses (IAs) for
futility attractive: stopping a trial that is unlikely to succeed
saves resources and avoids exposing patients to ineffective
treatments.

However, a futility IA at a low information fraction (e.g.\ 30%) has
limited statistical power. The data collected so far carry substantial
uncertainty, making it difficult to distinguish a truly futile
scenario from random noise. This is where **historical borrowing**
via a Meta-Analytic-Predictive (MAP) prior can help: by incorporating
external evidence on the control rate and overdispersion, the MAP
prior effectively supplements the interim data, increasing the
information fraction available at the IA and enabling more precise
futility decisions.

This article develops the idea in four steps:

1. **Baseline:** Operating characteristics (OC) of a fixed-sample
   design without an IA.
2. **Unblinded futility IA without borrowing:** Introduce a futility
   stop at 30% information fraction using non-informative
   priors. Quantify the power loss under the alternative.
3. **Unblinded futility IA with MAP prior:** Derive a MAP prior from
   four historical phase III trials. Show the effective sample size
   (ESS) and the information fraction gained.
4. **Interim decision-making:** Conditional power and predictive power
   (probability of success) at a hypothesized interim state.
5. **Design-level power:** Numerical integration over the full trial
   to quantify the expected power loss from the futility IA under the
   alternative.

Throughout, the **final confirmatory analysis is purely frequentist**
— the MAP prior is used only for interim projections.


# Negative binomial setup

In clinical trials with count data, where more than one event can
happen for a patient — such as asthma exacerbations — the **negative
binomial** distribution is often used. One of several parametrizations
uses a mean rate and a dispersion parameter. It extends the Poisson
distribution — which assumes that all patients with the same covariate
values share a single event rate — by allowing the event rate to vary
between patients (the larger the dispersion parameter, the more the
event rate varies between patients). Differences in follow-up due to
administrative censoring can be reflected using a **log-follow-up
offset**, i.e. the expected number of events for a patient is their
event rate times their follow-up time.

We work on the **log mean rate** scale under the standard normal
approximation to the MLE (see [Appendix](#appendix) for the full
parametrization). The treatment effect is the log-rate ratio
$\delta = \log(\lambda_T / \lambda_C)$.

```{r nb-setup}
## Assumed true parameters for design planning
lambda_ctrl  <- 1.8          # placebo rate (events/patient-year)
kappa_true   <- 1.9          # overdispersion (phi)
## Effective average follow-up per patient (years). Rather than a full
## year for everyone, we allow for ~15% dropout with, on average, half
## the intended follow-up (cf. Holzhauer et al., 2018):
##   1 * 0.85 + 0.5 * 0.15 = 0.925 patient-years.
followup     <- 1 * 0.85 + 0.5 * 0.15
log_exposure <- log(followup)

## NB family object for OC calculations (fixed kappa)
nb_family <- negative.binomial(theta = 1 / kappa_true)

## Per-patient sampling SD on the log-rate scale
## sigma = sqrt((1 + kappa * lambda * t) / (lambda * t))
sigma_fn <- function(lambda, kappa, t) {
  sqrt((1 + kappa * lambda * t) / (lambda * t))
}
sigma_ctrl  <- round(sigma_fn(lambda_ctrl, kappa_true, followup), 2)
sigma_treat <- sigma_ctrl  # same rate assumed under H0

cat(sprintf("Per-patient sigma (log-rate scale): %.3f\n", sigma_ctrl))
```


# Fixed-sample design (no interim)

We first evaluate the operating characteristics of a conventional
2-arm trial without an interim analysis.

```{r fixed-design}
## Priors — non-informative for both arms (no borrowing)
uninf_ctrl  <- mixnorm(c(1, log(lambda_ctrl), 10), sigma = sigma_ctrl)
uninf_treat <- mixnorm(c(1, log(lambda_ctrl), 10), sigma = sigma_treat)

## Sample sizes
n_treat <- 225
n_ctrl  <- 225

## Decision: P(delta < 0 | data) > 0.975  (one-sided test)
success_crit <- decision2S(0.975, 0, lower.tail = TRUE)

## OC function
oc_fixed <- oc2S(uninf_treat, uninf_ctrl,
  n1 = n_treat, n2 = n_ctrl,
  decision = success_crit,
  family = nb_family,
  offset1 = log_exposure, offset2 = log_exposure
)
```

```{r fixed-power, fig.cap="Power curve — fixed-sample design (no IA)", echo = FALSE}
log_mu_ctrl <- log(lambda_ctrl)
log_rr <- seq(0, -0.8, by = -0.025)

power_fixed <- oc_fixed(log_mu_ctrl + log_rr,
                        rep(log_mu_ctrl, length(log_rr)))

power_df <- data.frame(log_rr = log_rr, power = power_fixed)

## The sample size (n = 225/arm) is chosen so that 80% power is reached
## exactly at the design alternative (rate ratio 0.65); mark this point.
rr_alt <- 0.65

ggplot(power_df, aes(exp(log_rr), power)) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = 0.8, linetype = "dashed", colour = "grey50") +
  geom_vline(xintercept = rr_alt, linetype = "dotted", colour = "steelblue") +
  geom_point(data = data.frame(x = rr_alt, y = 0.8),
             aes(x, y), colour = "steelblue", size = 2) +
  annotate("text", x = rr_alt * 0.99, y = 0.5,
           label = "design alternative\n(80% power)",
           hjust = 1, size = 3, colour = "steelblue") +
  labs(x = "Rate ratio (treatment vs control)",
       y = "Power",
       title = "Power curve — no interim analysis") +
  scale_x_log10(breaks = c(0.5, rr_alt, 0.8, 1.0)) +
  scale_y_continuous(breaks = c(0, 0.25, 0.5, 0.75, 0.8, 1.0))
```


# Futility IA at 30% information fraction (no borrowing)

We now add a futility stop at 30% of the planned information. With no
historical borrowing, the IA relies entirely on the accrued trial
data. This would be a very early stop with rather imprecise estimates
compared to a more common IA look at 50%. 

As a simplifying assumption we treat the interim analysis as if it
were based on 30% of the patients with complete follow-up. In practice
the same information would come from more patients, many of them with
only partial follow-up at the time of the interim look; their data can
still be used through the use of an offset in the regression. This distinction is
not merely cosmetic: for a recurrent-event endpoint the information is
event-driven, so a given amount of *total* follow-up
($N \times$ expected follow-up per patient) is worth slightly more when
it comes from many patients observed for a short time than from fewer
patients each observed to completion. The calculation here is thus
somewhat simplfying the actual situation.

The information fraction is defined as the ratio of the Fisher
information for the treatment effect (log rate ratio) at the interim
relative to the final analysis. Under a negative binomial model with
log link, the per-patient Fisher information for $\log\mu$ in arm $j$
is $I_j = \lambda_j t / (1 + \kappa\lambda_j t)$, and the information
for the treatment effect $\delta = \log\mu_T - \log\mu_C$ from $n_T$
and $n_C$ patients is
$$
\mathcal{I}(\delta) = \frac{1}{(n_T I_T)^{-1} + (n_C I_C)^{-1}}.
$$
We compute the interim sample sizes required for 30\% of
$\mathcal{I}(\delta)$ at the planned final sample size, evaluated
under the design alternative $\log(\text{RR}) = \log(0.65) \approx -0.43$:

```{r ia-setup, echo=FALSE}
## Design alternative: log rate ratio at which the trial is evaluated
## (rate ratio 0.65; the sample size n = 225/arm is chosen so that the
## fixed design reaches 80% power exactly at this alternative)
log_rr_design        <- log(0.65)
lambda_treat_design  <- lambda_ctrl * exp(log_rr_design)

## Per-patient Fisher information for log(mu) under each arm
## In a NB GLM with log link: I_i = lambda*t / (1 + kappa*lambda*t)
fisher_per_patient <- function(lambda, kappa, t) {
  lambda * t / (1 + kappa * lambda * t)
}

Ip_ctrl  <- fisher_per_patient(lambda_ctrl, kappa_true, followup)
Ip_treat <- fisher_per_patient(lambda_treat_design, kappa_true, followup)

## Fisher information for delta at planned sample size
info_final  <- 1 / (1 / (n_treat * Ip_treat) + 1 / (n_ctrl * Ip_ctrl))

## Target: 30% of final information at the IA
info_frac_ia <- 0.30
info_target  <- info_frac_ia * info_final

## Solve for interim n assuming 1:1 allocation preserved
n_ia_exact  <- info_target * (1 / Ip_treat + 1 / Ip_ctrl)
n_ctrl_ia   <- round(n_ia_exact)
n_treat_ia  <- round(n_ia_exact)

knitr::kable(data.frame(
  Quantity = c("Design alternative log(RR)",
               "Per-patient Fisher info (treatment)",
               "Per-patient Fisher info (control)",
               "Information at final (n/arm)",
               sprintf("Information target at IA (%.0f%%)", 100 * info_frac_ia),
               "Patients at IA (treatment / control)",
               "Patients remaining (treatment / control)"),
  Value = c(sprintf("%.2f", log_rr_design),
            sprintf("%.4f", Ip_treat),
            sprintf("%.4f", Ip_ctrl),
            sprintf("%.2f (n = %d)", info_final, n_ctrl),
            sprintf("%.2f", info_target),
            sprintf("%d / %d", n_treat_ia, n_ctrl_ia),
            sprintf("%d / %d", n_treat - n_treat_ia, n_ctrl - n_ctrl_ia))
))
```

The futility rule is based on **predictive power (probability of
success)**: given the interim posterior, what is the probability that
the final analysis — using all planned patients — will declare
success?

If the PoS falls below a futility threshold $\gamma_F$ (e.g.\ 10%),
we stop:

```{r futility-rule}
## Futility threshold for PoS
post_futility_thresh <- 0.10
```

## Operating characteristics of the IA design

To evaluate the operating characteristics of the design *with* the
futility IA, we use numerical integration via `oc2S_interim()`.  This
function avoids noisy Monte Carlo estimation via Gauss-Hermite
quadrature over the distribution of possible interim outcomes (see
[Appendix: Integration approach](#appendix-integration) for details).

The interim decision rule continues the trial whenever the
probability of success exceeds the futility threshold:

```{r oc-ia-noprior, fig.cap="Power with and without futility IA (no borrowing)"}
## IA continuation rule: continue if PoS > threshold.
## The predictive power (PoS) is computed from the interim projection
## posteriors: the (uninformative) treatment posterior and the control
## posterior. This exact same rule is reused below for the MAP design —
## only the control interim prior (prior2_ia) changes, not the rule.
## Note: we use the fixed-sigma path of pos2S (no family argument)
## for performance. This is valid because sigma is approximately
## constant over the posterior range at the interim.
ia_rule_pos <- function(post1_ia, post2_ia, post1_ia_info, post2_ia_info) {
  pos2S(post1_ia, post2_ia,
        n1 = n_treat - n_treat_ia, n2 = n_ctrl - n_ctrl_ia,
        decision = success_crit,
        sigma1 = sigma_ctrl, sigma2 = sigma_ctrl)(post1_ia_info, post2_ia_info) > post_futility_thresh
}

## OC function with futility IA (no borrowing).
## prior1/prior2 are uninformative (used for the final analysis boundary);
oc_ia_uninf <- oc2S_interim(
  prior1 = uninf_treat, prior2 = uninf_ctrl,
  n1 = n_treat, n2 = n_ctrl,
  n1_ia = n_treat_ia, n2_ia = n_ctrl_ia,
  decision = success_crit,
  ia_rule = ia_rule_pos,
  prior1_ia = uninf_treat, prior2_ia = uninf_ctrl,
  family = nb_family,
  offset1 = log_exposure, offset2 = log_exposure,
  Ngrid_ia = 31L
)

## Evaluate at null and the design alternative (pos2S rule is expensive)
log_rr_grid <- c(`No benefit` = 0, `Design alt` = log_rr_design)

power_ia_uninf_df <- oc_ia_uninf(
  theta1 = log_mu_ctrl + log_rr_grid,
  theta2 = rep(log_mu_ctrl, length(log_rr_grid))
)
power_ia_uninf_df$log_rr <- log_rr_grid 
```

```{r power-comparison-noprior}
## Compare with fixed-design power at same grid points
power_fixed_grid <- oc_fixed(log_mu_ctrl + log_rr_grid,
                             rep(log_mu_ctrl, length(log_rr_grid)))
```

```{r power-comparison-noprior-tab, echo=FALSE}
knitr::kable(data.frame(
  Scenario = names(log_rr_grid),
  `Rate ratio` = round(exp(log_rr_grid), 2),
  `Fixed design` = power_fixed_grid,
  `IA (no borrow)` = power_ia_uninf_df$power,
  `Power loss` = power_fixed_grid - power_ia_uninf_df$power,
  `Futility stop` = power_ia_uninf_df$stop_prob,
  check.names = FALSE
), digits = 3, row.names = FALSE,
caption = "Power and futility-stop probability by scenario (no borrowing, PoS rule)")
```

At 30% information fraction with non-informative priors, the futility
IA already behaves quite well: it stops a clearly ineffective drug
(rate ratio = 1) with high probability — well above one half — while
costing only about one and a half percentage points of power under the design
alternative. This is more effective than one might expect from such an
early look. What the non-informative design cannot do is separate
"stop for the right reason" from "stop because the 30%-information
estimate happened to look bad"; the interim data alone carry limited
information, so the decision is comparatively noisy. Historical
borrowing (next section) sharpens exactly this.


# MAP prior from historical phase III trials

## Historical data

We use placebo-arm data from the phase III asthma trials in the
`asthma` data set (Table 1 of Holzhauer, Wang & Schmidli, 2018):

```{r hist-data}
asthma_ph3 <- subset(asthma, phase == "phase III")
```

```{r hist-data-tab, echo=FALSE}
knitr::kable(asthma_ph3[, c("study", "d", "n", "log_mu_hat", "se_log_mu_hat", "kappa_hat")],
             digits = 2,
             caption = "Historical phase III placebo arms")
```

## MAP prior for the control log-rate

The historical trials have different exposure (follow-up) durations
$d$. Here the summaries provided in the `asthma` data set are already
on the yearly **log-rate** scale (log events per patient-year, back-calculated
from the raw counts and exposures), so `log_mu_hat` estimates
$\log\lambda_C$ directly and **no offset is needed**:

```{r map-rate}
map_mcmc <- gMAP(
  cbind(log_mu_hat, se_log_mu_hat) ~ 1 | study,
  ## NOTE: the historical inputs are already log-rates (per patient-year),
  ## so no exposure offset is used here. If instead the inputs were
  ## log-mean *counts* log(mu) = log(lambda * d) for exposure d, we would
  ## add `offset = log(d)` to recover the log-rate log(lambda). Whether an
  ## offset is needed therefore depends on how the historical data are
  ## reported.
  data   = asthma_ph3,
  family = gaussian,
  tau.dist  = "HalfNormal",
  tau.prior = sigma_ctrl / 4,
  beta.prior = 2
)

print(map_mcmc)
map_rate <- automixfit(map_mcmc)
sigma(map_rate) <- sigma_ctrl
```

```{r map-forest, fig.cap="Forest plot of MAP model for control event rate"}
pl <- plot(map_mcmc)

## The gMAP model works on the log-rate scale, so the estimates are
## spaced logarithmically. We keep that geometry but label the ticks
## with the natural event rate (events/patient-year) for readability.
## Because plot.gMAP uses coord_flip(), the estimate lives on the y
## aesthetic, so we override scale_y_continuous().
rate_breaks <- c(1.25, 1.5, 1.75, 2.0, 2.5, 3.0)
print(pl$forest_model +
  ggplot2::scale_y_continuous(
    name   = "Control event rate [events/patient-year]",
    breaks = log(rate_breaks),
    labels = rate_breaks))
```

## Effective sample size and information fraction gain

The ESS tells us how many patients the MAP prior is worth. It adds
precision to the **control arm only**, so we fold it into the
treatment-effect information $\mathcal{I}(\delta)$ defined above as
$\text{ESS}$ extra control patients. The information fraction at the IA
then uses the same $\mathcal{I}(\delta)/\mathcal{I}(\delta)_{\text{final}}$
ratio, now with $n_C \to n_{C,\text{IA}} + \text{ESS}$:

```{r ess-info}
ess_map <- round(ess(map_rate, sigma = sigma_ctrl))

## Effect information I(delta) at IA without borrowing
info_ia_no_borrow <- 1 / (1 / (n_treat_ia * Ip_treat) +
                          1 / (n_ctrl_ia * Ip_ctrl))
info_frac_no_borrow <- info_ia_no_borrow / info_final

## Effect information I(delta) at IA with MAP prior:
## ESS adds patient-equivalents to the CONTROL arm only.
info_ia_with_map <- 1 / (1 / (n_treat_ia * Ip_treat) +
                         1 / ((n_ctrl_ia + ess_map) * Ip_ctrl))
info_frac_with_map <- info_ia_with_map / info_final
```

```{r ess-info-tab, echo=FALSE}
knitr::kable(data.frame(
  Quantity = c("MAP prior ESS",
               "Info fraction at IA (no borrowing)",
               "Info fraction at IA (with MAP prior)",
               "Info fraction gained"),
  Value = c(sprintf("%d patients", ess_map),
            sprintf("%.1f%%", 100 * info_frac_no_borrow),
            sprintf("%.1f%%", 100 * info_frac_with_map),
            sprintf("+%.1f pp", 100 * (info_frac_with_map - info_frac_no_borrow)))
))
```

The MAP prior substantially increases the effective information
fraction at the IA, which we would expect to make futility decisions
more reliable.

## Power with futility IA and MAP prior

```{r oc-ia-map}
## Same PoS futility rule as before (ia_rule_pos): continue if
## PoS > threshold. The interim projection now uses the MAP-informed
## control posterior (post2_ia_info) via prior2_ia = map_rate below;
## the continuation rule itself is unchanged. The pos2S boundary uses
## the uninformative posteriors, consistent with an uninformative
## final analysis.

## OC function with futility IA and MAP prior
## prior1/prior2 are uninformative (used for the final analysis boundary).
## prior1_ia is the uninformative treatment prior and prior2_ia is
## the MAP prior; both are used only for the interim decision.
oc_ia_map <- oc2S_interim(
  prior1 = uninf_treat, prior2 = uninf_ctrl,
  n1 = n_treat, n2 = n_ctrl,
  n1_ia = n_treat_ia, n2_ia = n_ctrl_ia,
  decision = success_crit,
  ia_rule = ia_rule_pos,
  prior1_ia = uninf_treat, prior2_ia = map_rate,
  family = nb_family,
  offset1 = log_exposure, offset2 = log_exposure,
  Ngrid_ia = 31L
)

## Evaluate at two scenarios: no benefit (rate ratio = 1) and the
## design alternative. The pos2S-based rule is expensive, so we keep
## the grid small.
scenarios   <- c(`No benefit` = 0, `Design alt` = log_rr_design)
theta1_eval <- log_mu_ctrl + scenarios
theta2_eval <- rep(log_mu_ctrl, length(scenarios))

map_grid    <- oc_ia_map(theta1 = theta1_eval, theta2 = theta2_eval)
uninf_grid  <- oc_ia_uninf(theta1 = theta1_eval, theta2 = theta2_eval)
fixed_power <- oc_fixed(theta1_eval, theta2_eval)
```

```{r oc-ia-map-tab, echo=FALSE}
knitr::kable(data.frame(
  Scenario = names(scenarios),
  `Rate ratio` = round(exp(scenarios), 2),
  `Power (no IA)` = fixed_power,
  `Power (no borrow)` = uninf_grid$power,
  `Power (MAP)` = map_grid$power,
  `Stop (no borrow)` = uninf_grid$stop_prob,
  `Stop (MAP)` = map_grid$stop_prob,
  check.names = FALSE
), digits = 3, row.names = FALSE,
caption = "Power and futility-stop probability by scenario")
```

The table compares the two designs at a **common** PoS futility
threshold of 10%. Under the **design alternative** the futility IA
costs about 1.6 percentage points of power without borrowing
(80.0% → 78.4%); with the MAP prior the loss shrinks to about 1.0
percentage points (80.0% → 79.1%), and the chance of an erroneous
futility stop under the alternative falls (roughly 4.6% → 2.9%). Under
**no treatment benefit** (rate ratio = 1), however, the picture at this
shared threshold is mixed: the non-informative design actually stops
*more* often (about 63%) than the MAP design (about 57%). This is not a
contradiction — at a common PoS threshold the two rules are simply not
equally aggressive, so reading the null-stopping probabilities side by
side compares apples to oranges.

# Calibrating the futility threshold to a fixed power loss

Comparing the two designs at a *shared* futility threshold mixes two
effects: the different threshold behaviour of the two analyses in
terms of operating charachersitics and the different information they
carry. An alternative, as poroposed by Gallo, Mao & Shih (2014), fixes
one (Frequentist) operating characteristic for both methods and reads
off the others. A futility rule can be expressed equivalently on
several one-to-one scales — a test statistic, the observed effect
estimate, conditional power, or predictive power — so the choice of
scale is a matter of convenience once we agree on the operating
characteristics we care about. Here we **calibrate each method to the
same power loss of 2% under the design alternative** and then compare
the futility stopping probability under the null.

**A fast, equivalent futility scale.** Evaluating the PoS at every
quadrature node is expensive because each evaluation solves a small
predictive-power (double) integral. Since the futility scales are
one-to-one for a fixed design, we can instead use the interim
**posterior assurance** for a positive treatment effect — the
posterior probability that the treatment is better than control,
$P(\delta < 0 \mid \text{interim data})$, a monotone function of the
interim posterior mean difference. This is far cheaper to compute (no
inner predictive integral) yet traces the *same* power-loss /
stopping-probability trade-off as the PoS rule. Importantly, the
borrowing is fully retained: the control arm's interim posterior is
still formed from the MAP prior, so the assurance rule uses exactly
the same MAP-informed control estimate — it only changes the *scale*
on which the threshold is expressed, not the information used.

**A deliberate asymmetry.** Note that this uses the MAP prior *only for
the interim decision making*. The final confirmatory analysis remains purely
frequentist and never sees this prior. The interim decision analysis
and the final analysis are therefore intentionally different: borrowing
is a design/monitoring tool here, not part of the primary inference.

```{r assurance-rule}
## Fast, equivalent futility scale: continue iff the interim posterior
## assurance P(delta < 0) exceeds a threshold gamma. This is a monotone
## re-scaling of the PoS rule (Gallo, Mao & Shih, 2014) but avoids the
## inner predictive integral, so the threshold search is ~25x faster.
## The MAP prior still enters via the control interim posterior
## (prior2_ia = map_rate below).
assurance_rule <- function(gamma) {
  function(post1_ia, post2_ia, post1_ia_info, post2_ia_info) {
    pmixdiff(post1_ia_info, post2_ia_info, 0) > gamma
  }
}

make_oc_assurance <- function(gamma, prior2_ia, Ngrid_ia = 31L) {
  oc2S_interim(
    prior1 = uninf_treat, prior2 = uninf_ctrl,
    n1 = n_treat, n2 = n_ctrl,
    n1_ia = n_treat_ia, n2_ia = n_ctrl_ia,
    decision = success_crit,
    ia_rule = assurance_rule(gamma),
    prior1_ia = uninf_treat, prior2_ia = prior2_ia,
    family = nb_family,
    offset1 = log_exposure, offset2 = log_exposure,
    Ngrid_ia = Ngrid_ia
  )
}
```

The threshold that comes closest to a 2% power loss is found once with
`uniroot`. Because it depends only on the fixed design inputs, we
determine it in the (non-evaluated) block below and hard-code the
resulting values; they only change if the design inputs change.

```{r calibrate-thresholds, eval=FALSE}
## One-off calibration (not evaluated on render). Solves for the
## assurance threshold giving a 2% power loss under the design
## alternative, separately for each method.
## Design inputs these thresholds depend on:
##   lambda_ctrl = 1.8, kappa_true = 1.9, followup = 0.925,
##   n_treat = n_ctrl = 225, n_ia = 68/arm, log_rr_design = log(0.65),
##   MAP prior = map_rate (asthma phase III, no offset),
##   uninf_treat = N(0, 10), success_crit = P(delta<0) > 0.975.
target_power_loss <- 0.02
theta1_alt <- log_mu_ctrl + log_rr_design
theta2_alt <- log_mu_ctrl
power_alt_fixed <- oc_fixed(theta1_alt, theta2_alt)

## Two-stage search for speed. The interim OC is integrated by
## Gauss-Hermite quadrature, whose cost grows with Ngrid_ia. We first
## locate the threshold on a cheap coarse grid (Ngrid_ia = 11) over the
## full range, then refine on the accurate grid (Ngrid_ia = 31) seeded
## from a narrow bracket around the coarse solution. The bracket
## auto-extends (extendInt = "upX", as the power loss increases with
## gamma) so the refinement stays robust even though the coarse and
## accurate grids can disagree by more than the initial half-width.
solve_gamma <- function(prior2_ia) {
  gfun <- function(gamma, Ngrid_ia) {
    p <- make_oc_assurance(gamma, prior2_ia, Ngrid_ia)(theta1_alt, theta2_alt)["power"]
    (power_alt_fixed - as.numeric(p)) - target_power_loss
  }
  ## Stage 1: coarse, wide bracket.
  g_coarse <- uniroot(gfun, interval = c(0.05, 0.90), tol = 1e-3,
                      Ngrid_ia = 11L)$root
  ## Stage 2: accurate grid, narrow self-extending bracket around it.
  uniroot(gfun, interval = c(g_coarse - 0.01, g_coarse + 0.01),
          extendInt = "upX", tol = 1e-3, Ngrid_ia = 31L)$root
}

gamma_uninf <- solve_gamma(uninf_ctrl)  # -> 0.481
gamma_map   <- solve_gamma(map_rate)    # -> 0.691
```

```{r calibrated-thresholds}
## Calibrated assurance thresholds (targeting a 2% power loss under the
## alternative), from the one-off search above for the design inputs
## listed there. Note: the interim OC is integrated on a discrete
## Gauss-Hermite grid (Ngrid_ia = 31), so the achievable power loss is
## quantized in small steps and 2% is met only up to that granularity.
## These thresholds are the ones closest to a 2% loss (about 1.9% for
## the non-informative and about 2.1% for the MAP design); nudging gamma
## to the next grid step would jump the loss to ~2.6% / ~2.2%.
gamma_uninf <- 0.481
gamma_map   <- 0.691
```

With the thresholds fixed, we compare the two methods at their
respective calibrated operating points: the power under the alternative
(equal by construction, up to the ~2% loss and the quadrature
granularity noted above) and the futility stopping probability under
the null.

```{r calibration-grid}
calib_scen <- c(`No benefit` = 0, `Design alt` = log_rr_design)
ct1 <- log_mu_ctrl + calib_scen
ct2 <- rep(log_mu_ctrl, length(calib_scen))

oc_cal_uninf <- make_oc_assurance(gamma_uninf, uninf_ctrl)(theta1 = ct1, theta2 = ct2)
oc_cal_map   <- make_oc_assurance(gamma_map,   map_rate)(theta1 = ct1, theta2 = ct2)
fixed_cal    <- oc_fixed(ct1, ct2)
```

```{r calibration-tab, echo=FALSE}
knitr::kable(data.frame(
  Scenario = names(calib_scen),
  `Rate ratio` = round(exp(calib_scen), 2),
  `Power (no IA)` = fixed_cal,
  `Power (no borrow)` = oc_cal_uninf$power,
  `Power (MAP)` = oc_cal_map$power,
  `Stop (no borrow)` = oc_cal_uninf$stop_prob,
  `Stop (MAP)` = oc_cal_map$stop_prob,
  check.names = FALSE
), digits = 3, row.names = FALSE,
caption = "Both methods calibrated to a 2% power loss under the design alternative")
```

Calibrated to the same (approximately 2%) power loss, the two methods
differ clearly in their futility stopping under the null: the MAP design
stops a truly ineffective drug substantially more often than the
non-informative design (here about 63% versus 42%). In other words, for
the *same* price in power, borrowing buys a markedly higher chance of
correctly stopping a futile trial — the benefit the shared-threshold
comparison understated.

# Conditional and predictive power at interim

At an actual IA, two related quantities inform the go/no-go
decision:

- **Conditional power (CP):** The probability of final success
  *given a specific assumed true treatment effect*. This is a
  frequentist concept — the unknown parameters are fixed.
- **Predictive power / Probability of Success (PoS):** The
  probability of final success *integrating over parameter
  uncertainty*. This is a Bayesian concept — the unknown parameters
  are drawn from the interim posterior.

These two quantities are closely linked rather than complementary:
predictive power is essentially conditional power averaged over the
interim posteriors. They answer different questions — "how likely is
success *if* the effect is $x$?" versus "how likely is success
*accounting for* our uncertainty about the effect?" — but they are not
independent pieces of information.

## Hypothesized interim state

We consider a scenario where at the IA (30% information fraction),
the observed data suggest a moderate treatment effect:

```{r interim-state}
## Observed interim summary statistics
ia_log_mu_ctrl  <- 0.70     # placebo MLE on log-rate scale
ia_se_mu_ctrl   <- sigma_fn(exp(0.70), kappa_true, followup) / sqrt(n_ctrl_ia)
ia_log_mu_treat <- 0.35     # treatment MLE (log RR ≈ -0.35)
ia_se_mu_treat  <- sigma_fn(exp(0.35), kappa_true, followup) / sqrt(n_treat_ia)
```

```{r interim-state-tab, echo=FALSE}
knitr::kable(data.frame(
  Quantity = c("Interim log-rate control",
               "Interim log-rate treatment",
               "Observed log RR"),
  Value = c(sprintf("%.2f", ia_log_mu_ctrl),
            sprintf("%.2f", ia_log_mu_treat),
            sprintf("%.2f", ia_log_mu_treat - ia_log_mu_ctrl)),
  SE = c(sprintf("%.3f", ia_se_mu_ctrl),
         sprintf("%.3f", ia_se_mu_treat),
         "")
))
```

## Interim posteriors

```{r interim-posteriors}
## Control posterior with MAP prior
post_ctrl_ia <- postmix(map_rate, m = ia_log_mu_ctrl, se = ia_se_mu_ctrl)

## Treatment posterior with the uninformative prior
post_treat_ia <- postmix(uninf_treat, m = ia_log_mu_treat, se = ia_se_mu_treat)
```

## Conditional power

Conditional power evaluates the probability of final success
*assuming the true treatment effect equals a specific value*.
We construct this using `oc2S` with the interim posteriors as
"priors" for the remaining data:

```{r conditional-power, fig.cap="Conditional power at interim"}
## OC function conditioned on interim posteriors
cp_fn <- oc2S(post_treat_ia, post_ctrl_ia,
  n1 = n_treat - n_treat_ia,
  n2 = n_ctrl  - n_ctrl_ia,
  decision = success_crit,
  family = nb_family,
  offset1 = log_exposure, offset2 = log_exposure
)

## Evaluate across true log-rate ratios
log_rr_cp <- seq(0, -0.8, by = -0.025)
cp_values <- cp_fn(log_mu_ctrl + log_rr_cp,
                   rep(log_mu_ctrl, length(log_rr_cp)))

ggplot(data.frame(log_rr = log_rr_cp, cp = cp_values),
       aes(exp(log_rr), cp)) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = c(0.5, 0.8),
             linetype = "dashed", colour = "grey50") +
  geom_vline(xintercept = exp(ia_log_mu_treat - ia_log_mu_ctrl),
             linetype = "dotted", colour = "steelblue") +
  annotate("text",
           x = exp(ia_log_mu_treat - ia_log_mu_ctrl) * 1.05,
           y = 0.95, label = "observed effect",
           hjust = 0, size = 3, colour = "steelblue") +
  labs(x = "Assumed true rate ratio",
       y = "Conditional power",
       title = "Conditional power at 30% IA") +
  scale_x_log10()
```

## Predictive power (Probability of Success)

Predictive power integrates over the interim posterior distribution of
the treatment effect, providing a single summary that accounts for
parameter uncertainty:

```{r predictive-power}
## PoS for remaining patients given interim posteriors
pos_fn <- pos2S(post_treat_ia, post_ctrl_ia,
  n1 = n_treat - n_treat_ia,
  n2 = n_ctrl  - n_ctrl_ia,
  decision = success_crit,
  family = nb_family,
  offset1 = log_exposure, offset2 = log_exposure
)

## Compute predictive power and conditional power at observed effect in one pass
pos_value <- pos_fn(post_treat_ia, post_ctrl_ia)
cp_at_obs <- cp_fn(ia_log_mu_treat, ia_log_mu_ctrl)
```

```{r predictive-power-tab, echo=FALSE}
knitr::kable(data.frame(
  Metric = c("Conditional power at observed effect",
             "Predictive power (PoS)"),
  Value = c(sprintf("%.3f", cp_at_obs),
            sprintf("%.3f", pos_value))
))
```

The predictive power is typically lower than the conditional power
evaluated at the observed effect, because it accounts for the
possibility that the true effect is smaller (or absent). As a result
one would not usually use the same IA decision thresholds for
conditional and predictive power.


# Design-level power with futility IA

The individual-scenario analysis above is useful at the IA itself.
At the **design stage**, we want to know the overall impact of the
futility IA on power — averaging over all possible interim outcomes
under an assumed truth.

For Normal-endpoint group-sequential Bayesian designs, the
[gsbDesign](https://cran.r-project.org/package=gsbDesign) package
(Gerber & Gsponer, 2016) provides a full analytic treatment of such
operating characteristics via conjugate posterior updates. The
`oc2S_interim()` helper used here follows the same numerical
integration idea but works with the mixture-prior machinery in RBesT
(see [Appendix: Integration approach](#appendix-integration)).
We already computed these above. Calibrated to a common 2% power loss
under the design alternative (rate ratio = 0.65), the MAP prior
achieves a higher futility stopping probability under the null than the
non-informative design — exactly the desired behaviour, and the fair
comparison the reviewer-facing literature calls for.


# Discussion

This article demonstrated how historical information can be used only
for an interim futility analysis:

1. **Establish the baseline** without an IA.
2. **Quantify the cost** of adding a futility IA with limited data.
3. **Show the benefit** of a MAP prior, both in ESS/information
   fraction terms and in operating characteristics.
4. **Calibrate the decision threshold** to a common power loss and
   compare methods on an equal footing, examining the behaviour across
   a range of decision thresholds via conditional and predictive
   power.
5. **Confirm at the design level** that the power loss is acceptable
   via numerical integration.

Key insights:

- A futility IA at low information fractions (e.g.\ 30%) may be
  unreliable without supplementary information, because the uncertainty
  in the interim estimate is usually large.
- The MAP prior adds effective sample size to the control arm, boosting
  the information fraction at the IA and enabling sharper futility
  decisions.
- A useful way to read this benefit: borrowing lets us **"look into
  the future."** The historical information sharpens the *control* arm
  to a precision the trial would otherwise only reach at a later, more
  mature interim look (here the information fraction rises from 30% to
  about 44%). Because the treatment effect contrasts treatment against
  control, a sharper control estimate directly tightens the interim
  effect estimate, so we make the futility call as if from a
  higher-information look. This is a *partial* look into the future:
  only the control arm is advanced — the treatment arm still carries
  its actual interim information — and it reaches only as far as the
  (robust) MAP stays consistent with the concurrent control; under
  prior-data conflict the borrowing is down-weighted and the effective
  gain shrinks. Crucially, this is an *information* gain.
- The final confirmatory analysis remains frequentist — no prior is
  imported. The MAP prior serves only as a **decision supporting
  tool** for interim analyses.
- Predictive power (PoS) is, in a certain sense, a more conservative
  and realistic interim metric than conditional power evaluated at the
  observed effect, because it integrates over parameter uncertainty
  instead of conditioning on a single assumed effect. It is not
  uniformly more conservative — it can exceed the conditional power at a
  pessimistic assumed effect — but it avoids the over-optimism of
  plugging in the observed interim estimate.

Using historical information on the negative binomial dispersion
parameter can be done using a normal-normal hierarchical model and
then used as a MAP prior at the IA. However, uncertainty about the
overdispersion parameter is not supported with `RBesT` directly and
would require using e.g.\ `brms` instead.

## Limitations

- Overdispersion $\kappa$ is treated as fixed. A more thorough
  analysis would derive a MAP prior for $\kappa$ and integrate over
  its posterior uncertainty (cf.\ Holzhauer et al., 2018; see the
  [companion article on SSR](introduction_negbin_ia_ssr.html)).
- The normal approximation on the log-rate scale assumes moderate to
  large event counts. For rare-event settings, exact likelihoods may
  be needed.

## Acknowledgements

Many thanks to Björn Holzhauer for a thorough review of an earlier
version of this article.

### References

[1] Neuenschwander B. et al., *Clin Trials*. 2010; 7(1):5-18  
[2] Holzhauer B., Wang C., Schmidli H. Evidence synthesis from
    aggregate recurrent event data for clinical trial design and
    analysis. *Statistics in Medicine*. 2018;37:867-882.  
[3] Schmidli H. et al., *Biometrics* 2014;70(4):1023-1032  
[4] Gerber F, Gsponer T (2016). "gsbDesign: An R Package for Evaluating
    the Operating Characteristics of a Group Sequential Bayesian Design."
    *Journal of Statistical Software*, **69**(11), 1–23.
    doi: [10.18637/jss.v069.i11](https://doi.org/10.18637/jss.v069.i11).  
[5] Gsponer T, Gerber F, Bornkamp B, Ohlssen D, Vandemeulebroecke M,
    Schmidli H (2014). "A Practical Guide to Bayesian Group Sequential
    Designs." *Pharmaceutical Statistics*, **13**(1), 71–80.
    doi: [10.1002/pst.1593](https://doi.org/10.1002/pst.1593).  
[6] Gallo P, Mao L, Shih VH (2014). "Alternative Views on Setting
    Clinical Trial Futility Criteria." *Journal of Biopharmaceutical
    Statistics*, **24**(5), 976–993.
    doi: [10.1080/10543406.2014.932285](https://doi.org/10.1080/10543406.2014.932285).  

### R Session Info

```{r session, echo=FALSE}
sessionInfo()
```

```{r, include=FALSE}
options(.user_mc_options)
```

# Exercise: calibrate to a fixed null-stopping probability

The calibration above fixed the **power loss under the alternative** at
2% and then compared the two methods on their futility stopping
probability under the null. Because the futility scales are one-to-one
for a fixed design (Gallo, Mao & Shih, 2014), we can equally read the
trade-off the other way around: **fix the probability of killing a
truly ineffective drug under the null** and compare the resulting power
loss under the alternative. A sponsor who wants a guaranteed "kill
rate" for futile trials would prefer this dual calibration.

As an exercise, calibrate both methods to a common futility stopping
probability under the null — say 45% — and compare the power they
sacrifice under the design alternative. You should find the MAP prior
sacrifices *less* power for the same null-stopping guarantee, the
mirror image of the result in the calibration section.

```{r exercise-null-calibration, eval=FALSE}
## Dual calibration: fix the futility stopping probability under the
## null (H0: no treatment benefit) instead of the power loss under the
## alternative. Reuses assurance_rule() / make_oc_assurance() from the
## calibration section.
target_stop_null <- 0.45
theta_null <- rep(log_mu_ctrl, 2)         # control == treatment (RR = 1)
theta_alt  <- c(log_mu_ctrl + log_rr_design, log_mu_ctrl)

## Solve for the assurance threshold giving the target null-stop rate,
## using the same two-stage (coarse -> refined) search as the
## calibration section: locate on the cheap Ngrid_ia = 11 grid, then
## refine on Ngrid_ia = 31 with a self-extending bracket. The null-stop
## probability increases with gamma, so extendInt = "upX" applies.
solve_gamma_null <- function(prior2_ia) {
  gfun <- function(gamma, Ngrid_ia) {
    s <- make_oc_assurance(gamma, prior2_ia, Ngrid_ia)(theta1 = theta_null[1],
                                                       theta2 = theta_null[2])["stop_prob"]
    as.numeric(s) - target_stop_null
  }
  g_coarse <- uniroot(gfun, interval = c(0.05, 0.90), tol = 1e-3,
                      Ngrid_ia = 11L)$root
  uniroot(gfun, interval = c(g_coarse - 0.01, g_coarse + 0.01),
          extendInt = "upX", tol = 1e-3, Ngrid_ia = 31L)$root
}

g_uninf_null <- solve_gamma_null(uninf_ctrl)
g_map_null   <- solve_gamma_null(map_rate)

## Power under the alternative at each calibrated threshold, versus the
## fixed-design power, gives the power loss to compare across methods.
power_fixed_alt <- oc_fixed(theta_alt[1], theta_alt[2])
loss_uninf <- power_fixed_alt -
  make_oc_assurance(g_uninf_null, uninf_ctrl)(theta_alt[1], theta_alt[2])["power"]
loss_map   <- power_fixed_alt -
  make_oc_assurance(g_map_null, map_rate)(theta_alt[1], theta_alt[2])["power"]
## Expectation: loss_map < loss_uninf (MAP gives up less power for the
## same 45% chance of stopping a futile trial).
```

# Appendix: Integration approach for interim OC {#appendix-integration}

Computing the operating characteristics of a design with an interim
analysis requires averaging over all possible interim outcomes.
The `oc2S_interim()` function (available via
`system.file("extra", "oc2S_interim.R", package = "RBesT")`) does
this via **deterministic numerical integration** rather than Monte
Carlo simulation, yielding smooth, noise-free OC curves.

## Decomposition

The unconditional power is

$$
P(\text{success}) = E_{\bar{y}_{IA}}
\Big[\mathbf{1}\{\text{continue}(\bar{y}_{IA})\}
\cdot P(\text{success} \mid \bar{y}_{IA}, \theta)\Big]
$$

where $\bar{y}_{IA} = (\bar{y}_{1,IA}, \bar{y}_{2,IA})$ are the
interim summary statistics (MLEs on the log-rate scale) and $\theta =
(\theta_1, \theta_2)$ are the true parameters.

The outer expectation is a 2D integral over the joint distribution of
the interim summaries $\bar{y}_{IA} \sim N(\theta, \Sigma_{IA})$.
Since the two arms are independent, this factors into a product of
two 1D Normal integrals — handled by **Gauss-Hermite quadrature**.

## Full-trial boundary reuse

The key performance insight is that the **full-trial decision
boundary** $B(\bar{y}_2)$ can be computed once and reused at every
quadrature point.

The final analysis decides based on the full-sample means
$\bar{y}_{\text{full}}$, which are a deterministic function of the
interim and remaining data:

$$
\bar{y}_{j,\text{full}} =
\frac{n_{j,IA}\,\bar{y}_{j,IA} + n_{j,rem}\,\bar{y}_{j,rem}}{n_j}
$$

Conditional on the interim data $\bar{y}_{j,IA}$ and the true
parameter $\theta_j$, the full-sample mean is Normal:

$$
\bar{y}_{j,\text{full}} \mid \bar{y}_{j,IA},\theta_j \sim
N\!\left(
  \frac{n_{j,IA}\,\bar{y}_{j,IA} + n_{j,rem}\,\theta_j}{n_j},\;
  \frac{\sigma_j \sqrt{n_{j,rem}}}{n_j}
\right)
$$

The decision boundary $B(\bar{y}_{2,\text{full}})$ — computed by
`decision2S_boundary()` using the original priors and full sample
sizes — applies unchanged, because the sequential Bayesian update
property guarantees

$$
p(\theta \mid \bar{y}_{\text{full}})
= p(\theta \mid \bar{y}_{IA}, \bar{y}_{rem}).
$$

At each quadrature point, the conditional OC reduces to

$$
P(\text{success} \mid \bar{y}_{IA}, \theta) =
\int \Phi\!\Big(B(\bar{y}_{2,\text{full}}),\;
\mu_{1,\text{cond}},\; \text{se}_{1,\text{cond}}\Big)\;
\varphi(\bar{y}_{2,\text{full}};\,
\mu_{2,\text{cond}},\,\text{se}_{2,\text{cond}})\;
d\bar{y}_{2,\text{full}}
$$

which is again evaluated by Gauss-Hermite quadrature (an inner 1D integral).


# Appendix: Negative binomial parametrization {#appendix}

## Distribution

For patient $k$ in arm $j$ with follow-up (exposure) time $t_{jk}$
and mean event rate $\lambda_j$ (events per unit time), the count
$Y_{jk}$ follows

$$Y_{jk} \sim \text{NB}(\mu_{jk} = t_{jk}\,\lambda_j,\; \kappa)$$

with probability mass function

$$P(Y_{jk} = y) =
  \binom{y + \kappa^{-1} - 1}{y}\,
  \left(\frac{\kappa\,\mu_{jk}}{1 + \kappa\,\mu_{jk}}\right)^y\,
  \left(\frac{1}{1 + \kappa\,\mu_{jk}}\right)^{\kappa^{-1}}$$

and moments

$$E(Y_{jk}) = \mu_{jk} = t_{jk}\,\lambda_j, \qquad
  \text{Var}(Y_{jk}) = \mu_{jk} + \kappa\,\mu_{jk}^2
  = \mu_{jk}(1 + \kappa\,\mu_{jk}).$$

Here $\kappa \ge 0$ is the **overdispersion** parameter. For
$\kappa = 0$ the distribution reduces to Poisson. The treatment
effect is parametrized as the **rate ratio**
$\theta = \lambda_T/\lambda_C$, or equivalently the **log-rate ratio**
$\delta = \log\theta = \log\lambda_T - \log\lambda_C$.

## Parametrization conventions

Several conventions coexist in the literature:

| Convention | Overdispersion | Relation |
|------------|---------------|----------|
| $\kappa$ (this article, Holzhauer et al.) | $\kappa \ge 0$ | — |
| $\phi$ (Mütze et al., rpact) | $\phi \ge 0$ | $\phi = \kappa$ |
| `theta` in `MASS::negative.binomial()` | `theta` $> 0$ | `theta` $= 1/\kappa$ |
| `size` in `dnbinom()` | `size` $> 0$ | `size` $= 1/\kappa$ |

**Caution:** R's `MASS::negative.binomial(theta)` uses `theta` as
the *reciprocal* of the overdispersion ($1/\kappa$), so larger values
of `theta` mean *less* overdispersion.

## Fisher information on the log-rate scale

The analysis is conducted on the **log mean rate** scale,
$\nu_j = \log\lambda_j$. The MLE $\hat\nu_j$ is approximately
normally distributed with variance (Fisher information inverse)

$$\text{Var}(\hat\nu_j) \approx
  \frac{1 + \kappa\,\lambda_j\,t}{n_j\,\lambda_j\,t}$$

for $n_j$ patients each with exposure $t$. The approximate Fisher
information for arm $j$ is therefore

$$\mathcal{I}_j = \frac{n_j\,\lambda_j\,t}{1 + \kappa\,\lambda_j\,t}.$$

Key observations:

- **Precision is event-driven:** The numerator $n_j\lambda_j t$
  represents the expected total number of events. More events yield
  more information.
- **Overdispersion is a tax:** The denominator
  $1 + \kappa\lambda_j t$ inflates the variance. Higher $\kappa$
  reduces the information per patient.
- **Exposure matters:** Longer follow-up $t$ increases the expected
  events and thus the information, but with diminishing returns when
  $\kappa\lambda_j t \gg 1$.

The variance of the log-rate ratio estimator is

$$\text{Var}(\hat\delta) \approx
  \frac{1}{\mathcal{I}_T} + \frac{1}{\mathcal{I}_C}.$$

## Connecting to `MASS::negative.binomial()`

In RBesT's design functions (`oc1S`, `oc2S`, `pos1S`, `pos2S`), the
`family` argument accepts an object created by
`MASS::negative.binomial()`. This object encodes the variance–mean
relationship $V(\mu) = \mu + \mu^2/\texttt{theta}$, which on the
log-link scale translates exactly to the Fisher information formula
above.

```{r family-example}
## Create NB family with kappa = 1.9 (theta = 1/1.9)
kappa <- 1.9
nb <- MASS::negative.binomial(theta = 1 / kappa)

## The variance function: V(mu) = mu + mu^2 / theta = mu * (1 + kappa * mu)
mu_grid <- seq(0.5, 3, by = 0.5)
data.frame(
  mu     = mu_grid,
  V_mu   = nb$variance(mu_grid),
  check  = mu_grid * (1 + kappa * mu_grid)
)
```

```{r,include=FALSE}
options(.user_mc_options)
```
